---
title: "P8105 HW2"
author: "Ngoc Duong"
date: "10/03/2019"
output: github_document
editor_options: 
  chunk_output_type: console
---
### Problem 1
```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(knitr)
# read and clean Mr. Trash Wheel sheet
trash_data <- 
  read_excel("./subdirectory/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
            sheet = "Mr. Trash Wheel") %>%
            janitor::clean_names() %>%
            select(-c("x15","x16","x17")) %>%   # deselect variables that don't give useful information and have too many NA's
            drop_na() %>%                       # omit rows that do not include dumpster specific data
            mutate(       
              sports_balls = round(sports_balls, 0)) %>% # round sports_balls to nearest integers
            mutate(
              sports_balls = as.integer(sports_balls))   # convert sports_balls into an integer variable
```

```{r precipitation 2017 and 2018, message=FALSE, warning=FALSE}

#read and clean precipitation data for 2017 and 2018
precip_2018 <- 
  read_excel("./subdirectory/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
            sheet = "2018 Precipitation", skip = 1) %>%  # select appropriate sheet, load, and skip first header/row because it doesn't contain info 
            janitor::clean_names() %>%
            drop_na() %>%                       # omit rows without precipitation data 
            rename (precip_total = total) %>%   # rename precip_total variable 
            mutate(year = 2018)                 # add variable year for 2018

precip_2017 <- 
  read_excel("./subdirectory/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
            sheet = "2017 Precipitation", skip = 1) %>%  # select appropriate sheet, load, and skip first header/row because it doesn't contain info 
            janitor::clean_names() %>%          
            drop_na() %>%                       # omit rows without precipitation data 
            rename(precip_total = total) %>%    # rename precip_total variable 
            mutate(year = 2017)                 # add variable year for 2017

# Combine dataset and convert month to character variable
precip_merged = 
  left_join(
    precip_2017, precip_2018, by = "month") %>% # combine using left_join
  mutate(
    month = month.name[month],                  # turn month into a character variable using month.name[] 
    month = str_to_lower(month)) %>%            # make lower case strings (to standardize across datasets)
  rename (
    precip_2017 = precip_total.x,               # rename variable to avoid year confusion
    precip_2018 = precip_total.y) %>% 
  select(-c("year.x","year.y"))                 # drop year variables because not necessary and can be confusing 

```

Mr. Trash Wheel dataset has `r nrow(trash_data)` observations (rows) of `r ncol(trash_data)` variables. Key dumpster- and time-specific (identifying) variables include ``r names(trash_data)[1:4]``, and dumpster characteristic variables are ``r names(trash_data)[5:7]``, while the rest, such as ``r names(trash_data)[7:11]``, etc. show the amounts of each trash type in the dumpsters. More specifically, for example, median number of sports balls in a dumpster in 2017 was `r trash_data %>% filter(year == "2017") %>% summarize(median(sports_balls))` balls.

The 2017 preciptation dataset has `r count(precip_2017)` observations, and the 2018 precipitation dataset has `r count(precip_2018)` observations. Key variables for both precipitation datasets include `month`, `year`, and `precip_total` denoting total precipitation for each month. Total precipitation in 2018 (for the months with available data) was `r sum(pull(precip_merged, precip_2018), na.rm = TRUE)`, while total precipitation in 2017 was `r sum(pull(precip_merged, precip_2017), na.rm = TRUE)`. 

### Problem 2
```{r message=FALSE, warning=FALSE}
#import and clean pols dataset, arrange by year and month
pols_data <- read_csv("./subdirectory/pols-month.csv") %>% 
                  separate(mon, c("year","month","day")) %>%    # break up variable mon into integers variables year, month, and day
                  mutate(
                    month = as.numeric(month), 
                    year = as.numeric(year)) %>%                # make year and month numeric for month.abb[]
                  mutate(month = str_to_lower(month.abb[month]) # turn month into character using month.abb[] and standardize to lower case 
                    ) %>% 
                  arrange(year, match(month,month.abb)) %>%    # arrange year and month (using match to align character variable 'month' with pre-set month.abb)
                  mutate(
                    prez_gop = replace(prez_gop, prez_gop == 2, 1)) %>% # turn value 2 in prez_gop to 1 because they both indicate a republican president at that point in time 
                  mutate(
                    president =  prez_gop - prez_dem) %>%     # make numeric variable president combining data from prez_gop and prez_dem
                  mutate(
                    president = recode(president, `-1` = "dem", `1` = "gop")) %>% # recode president from numeric to character demonstrating dem or gop 
                  select(year, month, everything(), -c("prez_dem","prez_gop","day")) # make year and month leading columns and deselect unnecessary variables (captured somewhere else) 
```

```{r snp data, message=FALSE, warning=FALSE}
#import and clean snp data, arrange by year and month, and make them leading columns
snp_data <- read_csv("./subdirectory/snp.csv") %>% 
            separate(date, c("month","day","year")) %>%  # break up variable mon into integers variables year, month, and day
                  mutate(
                    month = as.numeric(month),       
                    year = as.numeric(year)) %>%         # make numeric to arrange 
                  arrange(year, month) %>%               # arrange year and month 
                  mutate(
                    month = str_to_lower(month.abb[month])) %>%  # # turn month into character using month.abb[] and standardize to lower case 
                  select(year, month, close, -day)       # select neccessary variables and rearrange so year and month are leading columns
```

```{r unemployment data, message=FALSE, warning=FALSE}
#import and clean snp data, arrange by year and month, and make them leading columns
unemployment_data <- read_csv("./subdirectory/unemployment.csv") %>% 
                  janitor::clean_names() %>% 
                  pivot_longer(jan:dec,      #switching from “wide” to “long” format using pivot_longer
                               values_to = "unemployment_rate",  
                               names_to = "month") %>%          
                  arrange(year, match(month,month.abb))  # arrange so data looks organized

# joining snp_data into pols_data using left_join, joining by year and month
aggregate_data <- left_join(
  pols_data, snp_data, by = c("year", "month"))

# joining unemployment_data into the pols-and-snp-aggregate data using left_join, joining by year and month
aggregate_data <- left_join(
  aggregate_data, unemployment_data, by = c("year","month")) %>%
                  arrange(year, match(month, month.abb)) # arrange year and month 
```

`unemployment_data`'s key variables (after cleaning) are ``r names(unemployment_data)``, which specify time points (monthly) across years and show the corresponding unemployment rate at each time point. `snp_data`'s key variables (after cleaning) include ``r names(snp_data)``,  which also specify time points (monthly) across years and show the corresponding closing Standard & Poor’s stock at each time point over the years.

`pols_data` also contains time-specific variables (`month` and `year`), and more importantly, the political affiliation of the President at each presented point in time, denoted by `president`. On top of that, `r names(pols_data)[3:8]` are variables that show the number of governors, senators, and representatives by democratic and republican affiliation at each (monthly) point in time over the years. 

The resulting dataset `aggregate_data` is a `r nrow(aggregate_data)` x `r ncol(aggregate_data)` dataframe. Variable `year`'s range is `r min(pull(aggregate_data,year))` - `r max(pull(aggregate_data,year))`, and names of key variables are ``r names(aggregate_data)[c(1:2,9:11)]``. 

It might also be worth noting that, `pols_data` contains months from 01/1947 to 06/2015, whereas `snp_data` covers months from 01/1950 to 07/2015 and `unemployment_data` covers 01/1948 to 06/2015, so there is a slight misalignment in time across these three datasets. Since we are merging `unemployment_data` and `snp_data` into `pols_data` (the "reference" dataset) using `left_join` and by `year` and `month`, this will result in missing data (NA's) in variables `close` (from `snp_data`) and `unemployment_rate` (from `unemployment_data`) during the period from 01/1947 to 01/1948 and 01/1950. Additionally, `close` datapoints after 06/2015 was dropped due to the same principle with left joining that's mentioned above.

### Problem 3
```{r message=FALSE, warning=FALSE}

baby_name_data <- read_csv("./subdirectory/Popular_Baby_Names.csv") %>% 
                  janitor::clean_names() %>% 
                  distinct() %>%        # use distinct() to get rid of duplicated rows
                  rename(name = childs_first_name, birthyear = year_of_birth) %>% # rename to avoid typing long strings
                  mutate(ethnicity = str_to_lower(ethnicity),  # standardize strings to lower case/to address inconsistent case structure more easily 
                         gender = str_to_lower(gender),
                         name = str_to_lower(name)) %>% 
                  mutate(ethnicity = as.factor(ethnicity),     
                         gender = as.factor(gender)) %>%      # turn ethnicity and gender into level/factor variables
                  mutate(ethnicity = 
                           recode(ethnicity, "asian and paci" = "asian and pacific islander",
                                  "black non hisp" = "black non hispanic",
                                  "white non hisp" = "white non hispanic")) %>% # rename duplicated factors/levels within ethnicity 
                  arrange(birthyear)   # arrange by year so data is organized
```
Side comment: I used summary(pull(baby_name_data,ethnicity)) to inspect the levels inside ethnicity variable before recode the levels.

```{r name tables, message = FALSE, warning = FALSE}
# table showing the rank in popularity of the name "Olivia"
baby_name_data %>% filter(name == "olivia") %>%        # filter with name "olivia" as criterion
                   select(-c("name","gender","count")) %>% # deslect name, gender (unnecessary) and count (unique values will mess up pivot_wider table output) 
                   pivot_wider(
                       names_from = "birthyear",       # show birthyear as variabl
                       values_from = "rank") %>% 
                   knitr::kable()                      # make table using kable 
```
Values in the cells are rank of popularity for the name Olivia 

```{r popular male name table, message = FALSE, warning = FALSE}
# Most popular male name table
baby_name_data %>% filter(gender == "male" & rank == 1) %>% # filter with male and rank 1 as criteria 
                   select(-c("count", "gender", "rank")) %>%                  # deselect count (whose unique/non-collapsible values will affect how pivot_wider works), gender and rank because they are implied when subsetting
                   pivot_wider(
                       names_from = "birthyear",            # take birthyear values to make variables (columns) and most popular names as values (observations)
                       values_from = "name") %>%              
                   knitr::kable()                           #make table using kable 
```

```{r male non hispanic white scatterplot, message=FALSE, warning=FALSE}
white_male_table <- baby_name_data %>% 
                    filter(gender == "male" &      
                           birthyear == 2016 & 
                           ethnicity == "white non hispanic") %>% # filter with male, white non-hispanic, and birthyear of 2016 as criteria 
                    arrange(count)                 # arrange count so dataset looks more organized

# scatterplot for number of children with (count on y axis) against popularity rank (x axis)
ggplot(white_male_table, aes(x= rank, y= count)) +
  geom_point(col = 4, alpha = .3) +
  labs(
    y="Count of names", 
    x = "Popularity rank from 1-100") + # adding name for x- and y-axis 
  ggtitle("Count of names and their rank in popularity for \nmale white non-hispanic children born in 2016")                 #adding title for scatterplot
```

